\chameleon{}: Untangling Type Errors Through Interactive Visualization and Exploration

### 2022-11-29 sections/related-work.tex
Wide images

@@

Another related topic is type-explanation~\cite{yang1999explaining, jun2002explaining}. Yang~\cite{jun2002explaining} showed an alternative type inference system capable of producing a human-like text explanation for why expressions are assigned certain types. A good explanation is drawn from surveying how human experts explain types. The resulting algorithm $\mathcal{H}$, generates a succinct explanation of the type inference steps to avoid using internal constructs (such as type variables). The explanation system has the advantage of acting like a human expert. However, when presented as text-based output, explanation systems have the potential to become verbose when types are complex or variable names are long. In \chameleon{}, we attempt to address this problem by showing one step of explanation at a time and referring to variables instead of spelling out the full name.

% Another method to improve type error message is search based type
% checking~\cite{lerner2007searchingtypeerror}. This method incrementally modifies
% an ever smaller part of the abstract syntax tree and replacing it with a
% wildcard expression and querying the type checker to verify if the program still
% type-checks. Lerner's study showed how two implementations could provide
% high-quality error messages and to suggest accurate fixes. While this system is
% great for suggesting solutions, is it limited in its ability to explain the type
% errors due to the fact that the system is unaware of any inner workings of the
% type systems.

% Some studies \cite{seidel2017learningtoblame} proposed a heuristic approach to
% improve type error messages. These studies use part of type information in the
% program to query external databases. Based on the external databases'
% capability, some projects can interactively fix the type error on the fly with
% reasonable accuracy. \pjs{Add jusdgment/comparison sentence!}

Debugging using a GUI Interactive Development Environment has been the standard practice for a very long time. Compared to a command line-based interface, a graphic user interface provides programming tools with the ability to show information hierarchy, provide immediate feedback to changes, and display complex visualizations. RxFiddle~\cite{banken2018debugging} is a tool to visualize data flow and value changes in a functional reactive program. Whyline~\cite{ko2009finding} is a Java debugging system that allows a user to ask questions like "why does variable X have value Y". It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error. Both  systems target programming runtime systems instead of types systems and static analysis. In fact, support for type systems has been lacking in both academic and industry software engineering tool design. The most common approach for supporting type systems we see in current programming editors and IDEs is to display type annotations when hovering on an expression and show a squiggly line to indicate the type error location. In \chameleon{}, colours and geometric shapes are used to create visual clarity of different components of a type error; interactivity is used to show conditional information and to change the granularity of details.

>>>>

Another direction to approach type error debugging is suggesting fixes. Sheng Chen's works [1,2] on counter factual typing is able to suggesting changes to make the ill-typed program typecheck. While this is promising ground, the work is not evaluated with human participants.  \cite{lerner2007searchingtypeerror} used an approach by repeatedly modifying the abstract syntax tree and querying the compiler/type checker to know whether the program still well-typed. This approach allows the algorithm to produce suggestions the syntax changes that could fix the type error. Benefiting from not needing a type system impelemntation and being language agnositic, the underlying idea was implemented and evaluated on 10 CS students. The system can be successfully adopted in not suitable for teaching and learning purpose, as it is not able to explain why the type error happened and why fixes are suggested. The lack of explanability is because the information was witheld by the type checker and the inference algorthm does not know (need to know) how types are computed.

One related approach is CEME (compiler error message enhancement). Prather showed that ECEM casts a positive result in understanding the type error through a series of mixed methods study. Decafe is a tool that is able to rephase the compier error message into an enhanced version. A study of over 200 CS1 students showed that CEME reduce overall errors in their assignments. Berik proposed a framework of constructing compiler error message based on the augumentation theory. In this proposal,  Berik argued that error messages  follow a simple argumentation layout or an exntended argumentation layout will result in a human friendly errors.  These research shows the significance of improving the language in a compiler error message. Most principles and suggestions are followed in \chameleon{} in concustructing error statements. However they were not targeting type errors but compiler error (some even include runtime error) in general. The nuances of type errors, such as alternative typings were not taken into account.

@@

% Another related topic is type-explanation~\cite{yang1999explaining, jun2002explaining}. Yang~\cite{jun2002explaining} showed an alternative type inference system capable of producing a human-like text explanation for why expressions are assigned certain types. A good explanation is drawn from surveying how human experts explain types. The resulting algorithm $\mathcal{H}$, generates a succinct explanation of the type inference steps to avoid using internal constructs (such as type variables). The explanation system has the advantage of acting like a human expert. However, when presented as text-based output, explanation systems have the potential to become verbose when types are complex or variable names are long. In \chameleon{}, we attempt to address this problem by showing one step of explanation at a time and referring to variables instead of spelling out the full name.

Debugging using a GUI Interactive Development Environment has been the standard practice for a very long time. Compared to a command line-based interface, a graphic user interface provides programming tools with the ability to show information hierarchy, provide immediate feedback to changes, and display complex visualizations. Provide opportunities to run consecutive command to query the information about code or executing command based on the existing code.
HazelTutor is a interactive filling type holes by suggesting template expressions (strategies as called by the authors). It also provide a cursor based type inspector that allows programmers to query the type of parts of the program.
Whyline~\cite{ko2009finding} is a Java debugging system that allows a user to ask questions like "why does variable X have value Y". It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error.  These debugging tools are great motivations for the development of \chameleon{}. However they focus on different aspect of debugging. Java Whyline mainly tackle the problem of unintended runtime behavior while HazelTutor specialized in interactive actions supporting type holes.
